---
title: "Math189_Homework4"
author: "Leah Seseri"
date: "2025-05-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1:

### Part 1:
```{r}
library(ISLR)
data("Auto")

mpg_median <- median(Auto$mpg)
mpg01 <- ifelse(Auto$mpg > mpg_median, 1, 0)
df <- data.frame(mpg01 = mpg01, Auto)
head(df)
```

### Part 2:
```{r}
# compare with cylinders
boxplot(cylinders ~ mpg01, data = df, main = "cylinders vs. mpg01", col = "lightblue")
```
```{r}
# compare with displacement
boxplot(displacement ~ mpg01, data = df, main = "displacement vs. mpg01", col = "lightgreen")
```

```{r}
# compare with horsepower
boxplot(horsepower ~ mpg01, data = df, main = "horsepower vs. mpg01", col = "lightcoral")
```

```{r}
# compare with weight
boxplot(weight ~ mpg01, data = df, main = "weight vs. mpg01", col = "lightyellow")
```

```{r}
# compare with acceleration
boxplot(acceleration ~ mpg01, data = df, main = "acceleration vs. mpg01", col = "lightpink")
```
```{r}
# compare with year
boxplot(year ~ mpg01, data = df, main = "year vs. mpg01", col = "lavender")
```
```{r}
# compare with origin
barplot(table(df$origin, df$mpg01),
        beside = TRUE, col = c("steelblue", "orange", "darkgreen"),
        main = "origin vs. mpg01")
```

Based on the boxplots I created to compare the different features with mpg01, the feature that seems to most likely be useful in predicting mpg01 is cylinders because there is a drastic difference between the cylinders for cars that have an mpg01 of 1 that is higher than the median compared to those with 0 and have a slower mpg. It appears that displacement, weight, and horsepower could also be useful in predicting mpg01 as it appears that cars that have a lower mpg tend to have higher displacement, weight, and horsepower than those with a higher mpg. Also, origin could have an impact on predicting mpg because based on the barplot, it seems that the some regions have more cars with lower mpg than other regions.

### Part 3:
```{r}
set.seed(123)

# training set of size 300 so test set to size 92
train_indices <- sample(1:nrow(df), 300)

train <- df[train_indices, ]
test <- df[-train_indices, ]
```

### Part 4:

### Part 5: