---
title: "Discussion Week 8"
output: html_document
date: "2025-05-18"
---

# Part 1 Nonlinear Decision Boundary

In this exercise, we explore a nonlinear decision boundary formed by an **ellipse**. 

## Classification with an Elliptical Boundary

We consider the ellipse defined by:
\[
\frac{(X_1 + 2)^2}{4} + \frac{(X_2 - 3)^2}{16} = 1
\]
This corresponds to an ellipse centered at \((-2, 3)\), with:

- Semi-minor axis length \(2\) along the \(X_1\)-direction

- Semi-major axis length \(4\) along the \(X_2\)-direction

### Visualizing the Boundary

Sketch the curve
\[
\frac{(X_1 + 2)^2}{4} + \frac{(X_2 - 3)^2}{16} = 1
\]

```{r}
# Define center and axis lengths
h <- -2     # center x
k <- 3      # center y
a <- 2      # semi-minor axis (horizontal, along X1)
b <- 4      # semi-major axis (vertical, along X2)

# Generate ellipse points parametrically
theta <- seq(0, 2 * pi, length.out = 100)
x_ellipse <- h + a * cos(theta)
y_ellipse <- k + b * sin(theta)

# Plot the ellipse
plot(x_ellipse, y_ellipse, type = "l", asp = 1, col = "black",
     xlab = "X1", ylab = "X2", xlim = c(-6, 2), ylim = c(-2, 8),
     main = "Ellipse Classifier Decision Boundary")

# Fill the ellipse
polygon(x_ellipse, y_ellipse, col = "lightgray", border = NA)
lines(x_ellipse, y_ellipse, lwd = 2)


```


The region where  
  \[
  \frac{(X_1 + 2)^2}{4} + \frac{(X_2 - 3)^2}{16} \leq 1
  \]  
  is the **interior** of the ellipse.
  
 The region where  
  \[
  \frac{(X_1 + 2)^2}{4} + \frac{(X_2 - 3)^2}{16} > 1
  \]  
  is the **exterior** of the ellipse.

## Classifying Observations

We define a classifier as follows:

- Assign to the **blue** class if  
  \[
  \frac{(X_1 + 2)^2}{4} + \frac{(X_2 - 3)^2}{16} > 1
  \]

- Assign to the **red** class if  
  \[
  \frac{(X_1 + 2)^2}{4} + \frac{(X_2 - 3)^2}{16} \leq 1
  \]
  
  Determine the class of each observation: (–2, 3), (0, 3) , (-4, 7), (-2,-1).
  
```{r}
# Observations to classify
points_to_check <- data.frame(
  X1 = c(-2, 0, -4, -2),
  X2 = c(3, 3, 7, -1)
)

# Compute decision function values
decision_values <- ((points_to_check$X1 - h)^2) / a^2 + ((points_to_check$X2 - k)^2) / b^2
classes <- ifelse(decision_values <= 1, "red", "blue")

classes
```
  
# Part 2 Support Vector Machines (SVMs) 


## Dataset Introduction: OJ (Orange Juice Purchases)

The `OJ` dataset comes from a supermarket marketing study that tracks customer purchases of orange juice.  
The main goal is to predict whether a customer buys the **Citrus Hill (CH)** brand or the **Minute Maid (MM)** brand based on various features related to price, promotions, and brand loyalty.

This dataset contains **1,070 observations**, each representing a single purchase event.  
The response variable is `Purchase`, which indicates whether the customer chose `"CH"` or `"MM"`.

Key predictor variables include:
- `PriceCH`, `PriceMM`: The price of Citrus Hill and Minute Maid at the time of purchase  
- `DiscCH`, `DiscMM`: The dollar discount applied to each brand  
- `LoyalCH`: A score reflecting the customer's loyalty to Citrus Hill  
- `SpecialCH`, `SpecialMM`: Indicators for whether the brand was on a special display  
- `PctDiscCH`, `PctDiscMM`: Discounts expressed as a percentage of the original price  
- `StoreID`, `STORE`: Identifiers for the store where the purchase was made  
- `WeekofPurchase`: The week number in which the purchase occurred


```{r}
library(ISLR2)
data(OJ)
head(OJ)
str(OJ)
attach(OJ)
```


### Split the Data into Training and Test Sets
```{r}
set.seed(1)
n<-length(Purchase)
train = sample(1 : n, size = 800) 
test = (-train)

##
trainset<-OJ[train,]
testset<-OJ[test,]
```

### Fit a Linear SVM with Cost = 0.01

```{r}
library(e1071)
svmfit <- svm( Purchase ~., data=OJ,subset = train, kernel = 'linear', cost = 0.01, scale = FALSE)
summary(svmfit)
```

### Evaluate Performance on Training and Test Sets

```{r}
ypred <- predict(svmfit, OJ)

train.error <- mean(ypred[train] != Purchase[train])
train.error
test.error<- mean(ypred[test] != Purchase[test])
test.error
```

### Tune the Cost Parameter for the Linear Kernel

```{r}
tune.out<- tune(svm, Purchase ~., data=OJ[train,], kernel = 'linear', ranges = list(cost = c(0.01,0.1,1,5,10)) ,scale=FALSE)

summary(tune.out)
tune.out$best.parameters
```
### Re-fit the Best Model from Tuning

```{r}
bestmod <- tune.out$best.model
summary(bestmod)

ypred <- predict(bestmod, OJ)

train.error <- mean(ypred[train] != Purchase[train])
train.error
test.error<- mean(ypred[test] != Purchase[test])
test.error
```


### Fit an SVM with Radial Kernel

```{r}
tune.gaussian <- tune(
  svm, 
  Purchase ~.,
  data=OJ[train,], 
  kernel = "radial",
  ranges = list(
    cost = c(0.1, 0.5, 1, 1.5, 2, 5),
    gamma = c(0.1, 0.5, 1, 1.5, 2)
  ),
  scale = FALSE
)

# Show tuning results
summary(tune.gaussian)

# Extract best model
bestmod.gaussian <- tune.gaussian$best.model
summary(bestmod.gaussian)




ypred <- predict(bestmod.gaussian, OJ)

train.error <- mean(ypred[train] != Purchase[train])
train.error
test.error<- mean(ypred[test] != Purchase[test])
test.error


```

### Fit an SVM with Polynomial Kernel (Degree 2)

```{r}
svmpfit <- svm( Purchase ~., data=OJ,subset = train, kernel = 'polynomial', degree=2, cost = 0.01,scale = FALSE)
summary(svmpfit)

ypred <- predict(svmpfit, OJ)

train.error <- mean(ypred[train] != Purchase[train])
train.error
test.error<- mean(ypred[test] != Purchase[test])
test.error


```


### Visualize the Linear SVM Decision Boundary with Two Predictors


We define `PriceDiff = PriceCH - PriceMM`.

```{r}

# Create PriceDiff variable
OJ$PriceDiff <- OJ$PriceCH - OJ$PriceMM

# Subset training data using two predictors and the target
data.train.2d <- OJ[train, c("Purchase", "LoyalCH", "PriceDiff")]
data.train.2d$LoyalCH <- as.numeric(data.train.2d$LoyalCH)
data.train.2d$PriceDiff <- as.numeric(data.train.2d$PriceDiff)
data.train.2d$Purchase <- as.factor(data.train.2d$Purchase)

# Fit linear SVM using only the two predictors
bestmod.linear <- svm(Purchase ~ LoyalCH + PriceDiff,
                      data = data.train.2d,
                      kernel = "linear",
                      cost = 0.1,     
                      scale = FALSE)

# Plot the decision boundary
plot(bestmod.linear, data.train.2d)

```

```{r}
### Plot Decision Boundary on Test Set 

# Ensure correct data types
data.test.2d <- OJ[test, c("Purchase", "LoyalCH", "PriceDiff")]
data.test.2d$LoyalCH <- as.numeric(data.test.2d$LoyalCH)
data.test.2d$PriceDiff <- as.numeric(data.test.2d$PriceDiff)
data.test.2d$Purchase <- as.factor(data.test.2d$Purchase)

# Predict classes on test set
test_preds <- predict(bestmod.linear, newdata = data.test.2d)

plot(bestmod.linear, data.test.2d)

```

| Symbol         | Meaning                                      |
|----------------|----------------------------------------------|
| **Red circle (o)**   | True class = `"CH"`, predicted correctly         |
| **Black circle (o)** | True class = `"MM"`, predicted correctly         |
| **Red X (x)**        | True class = `"CH"`, predicted incorrectly       |
| **Black X (x)**      | True class = `"MM"`, predicted incorrectly       |


### ROC Curves: Linear vs. Gaussian SVM on the OJ Dataset
```{r}
library(ROCR)

# Define a function for plotting ROC curves
rocplot <- function(pred, truth, ...) {
  pred_obj <- prediction(pred, truth)
  perf <- performance(pred_obj, "tpr", "fpr")
  plot(perf, ...)
}
```


```{r}
# Compute decision values from the linear SVM on training data
fitted.linear.train <- attributes(
  predict(bestmod.linear, OJ[train, ], decision.values = TRUE)
)$decision.values

# Compute decision values from the radial SVM on training data
fitted.gaussian.train <- attributes(
  predict(bestmod.gaussian, OJ[train, ], decision.values = TRUE)
)$decision.values

# Plot ROC curves for training data
par(mfrow = c(1, 2))
rocplot(-fitted.linear.train, OJ$Purchase[train], main = "Training Data", col = "red")
rocplot(-fitted.gaussian.train, OJ$Purchase[train], add = TRUE, col = "blue")
legend("bottomright", legend = c("Linear SVM", "Gaussian SVM"),
       col = c("red", "blue"), lty = 1, cex = 0.8)

# Compute decision values from the linear SVM on test data
fitted.linear.test <- attributes(
  predict(bestmod.linear, OJ[test,], decision.values = TRUE)
)$decision.values

# Compute decision values from the radial SVM on test data
fitted.gaussian.test <- attributes(
  predict(bestmod.gaussian, OJ[test, ], decision.values = TRUE)
)$decision.values

# Plot ROC curves for test data
rocplot(-fitted.linear.test, OJ$Purchase[test], main = "Test Data", col = "red")
rocplot(-fitted.gaussian.test,  OJ$Purchase[test], add = TRUE, col = "blue")
legend("bottomright", legend = c("Linear SVM", "Gaussian SVM"),
       col = c("red", "blue"), lty = 1, cex = 0.8)
```

